---
title: 'RAGE: Check Data'
author: "John Thompson"
date: "29 March 2023"
output: 
  html_document:
    theme: cerulean
---

```{r setup, include=FALSE}
library(tidyverse)
library(fs)

cache    <- "C:/Projects/RCourse/Masterclass/RAGE/data/cache"

trainDF <- readRDS(path(cache, "training.rds"))

knitr::opts_chunk$set(echo = FALSE, 
                      message=FALSE)

theme_set(theme_light())
```

## Background

Data for RAGE project. File GSE168753_processed_data.csv.gz downloaded from https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE168753  


## Expression by Patient

```{r}
trainDF %>%
  pivot_longer(-id, values_to = "expression", names_to = "gene") %>%
  group_by(id) %>%
  summarise( mn = mean(expression),
             sd = sd(expression) )
```

Plot of standard deviation against mean
```{r}
trainDF %>%
  pivot_longer(-id, values_to = "expression", names_to = "gene") %>%
  group_by(id) %>%
  summarise( mn = mean(expression),
             sd = sd(expression) ) %>%
  ggplot( aes(x=mn, y=sd)) +
  geom_point() +
  labs( x = "Mean Expression",
        y = "St Deviation Expression",
        title = "Each point represents a single patient")
```

## Expression by Gene

```{r}
trainDF %>%
  pivot_longer(-id, values_to = "expression", names_to = "gene") %>%
  group_by(gene) %>%
  summarise( mn = mean(expression),
             sd = sd(expression) )
```

Plot of standard deviation against mean
```{r}
trainDF %>%
  pivot_longer(-id, values_to = "expression", names_to = "gene") %>%
  group_by(gene) %>%
  summarise( mn = mean(expression),
             sd = sd(expression) ) %>%
  ggplot( aes(x=mn, y=sd)) +
  geom_point() +
  labs( x = "Mean Expression",
        y = "St Deviation Expression",
        title = "Each point represents a single gene")
```

## Issues

- One patient has a very low standard deviation and a mean near zero. Possible technical failure.  
- Means expression very variable across patients. Suggests the need for normalisation. Standard deviation also quite variable.  
- Expression across genes is unremarkable. Perhaps a negative correlation between mean and standard deviation.